Constructive Preference Elicitation by Setwise Max-margin Learning

### 2016-01-27 paper/ijcai16.tex
paper: reorder the second exp figure for clarity.
small fixes

\paragraph{Notation.} We use boldface letters $\vx$ to indicate vectors,
uppercase letters $X$ for matrices, and calligraphic capital letters $\calX$
for sets. We frequently abbreviate the set $\{ x^i \}_{i=1}^n$ as $\{ x^i \}$
whenever the range of the index $i$ is clear from the context, and use $[n]$ as
a shorthand for the set $\{1, \ldots, n\}$. We write $\|\vx\|_1 := \sum_z |x_z|$
to indicate the $\ell_1$ vector norm, $\langle \cdot, \cdot \rangle$ to
denote the usual dot product, and $X'$ to denote matrix transposition.

@@

During learning, the actual weight vector $\vw$ is {\em unknown} to the
learning system, and must be estimated from pairwise choices elicited from the
user.

%\andrea{@PAOLO: qualche parola e reference per giustificare la scelta di pairwise queries?}  
Comparison queries are especially natural, asking a user if she prefers one option x to another y;
these comparisons can be extended to choice sets of more than two options \cite{viappiani2009,viappiani2010optimal}.
Comparison and choice queries are common in conjoint analysis \cite{louviere2000,toubia2004}.

@@

\section{Setwise Max-margin Learning}
\label{sec:formulation}

\paragraph{Non-linear Formulation.} We first introduce the problem
formulation as a mixed integer non-linear optimization problem, and
then show how to reduce it to a mixed integer linear optimization one.

The goal of our setwise max-margin approach is twofold. First, for any
given set size $k\geq 1$, we want to find a {\em set} of $k$ weight
vectors $\vw^{1}, \ldots, \vw^{k}$, chosen so that all user-provided
preferences are satisfied by the largest possible margin (i.e. all
weight vectors are consistent with respect to the user responses
$\calD$, modulo inconsistencies) and so that they are maximally
diverse.  Second, we want to construct a {\em set} of $k$
configurations $\vx^{1}, \ldots, \vx^{k}$, so that each configuration
$\vx^{i}$ is the ``best'' possref:keeney-mautible option when evaluated according to
the corresponding $\vw^{i}$ and configurations are maximally diverse
among each other. These options will be later employed to formulate
the user queries.

The first goal is achieved by translating all pairwise preferences
$\calD$ into ranking constraints: preferences of the form
$\vy^h_+ \succ \vy^h_-$ become linear inequalities of the form
$\langle \vw^i, \vy^h_+ - \vy^h_- \rangle \geq \mu$, where $\mu$ is the
{\em margin} variable (which we aim at maximizing) and $h$ ranges over
the responses.  Non-separable datasets, which occur in practice due to
occasional inconsistencies in user feedback, are handled by
introducing slack variables (whose sum we aim at minimizing). When
augmented with the slacks, the above inequalities take the form:
%
$$ \langle \vw^{i}, \vy^{h}_+ - \vy^{h}_- \rangle \ge \mu - \varepsilon^{i}_h $$
%
where $\varepsilon^{i}_h$ is the penalty incurred by weight vector $\vw^{i}$
for violating the margin separation of pair $h$. Indifference preferences, i.e.
$\vy^h_1 \approx \vy^h_2$, are translated as $|\langle \vw^i, \vy^h_1 - \vy^h_2 \rangle| < \varepsilon^i_h$;
the slack increases with the difference between the estimated utility of the
two options.

@@

\begin{figure}[b]
    \centering
    {\footnotesize
    \begin{tabular}{cccc}
        \hline
        \multicolumn{2}{c}{{\sc loss per iter.}} &
        \multicolumn{2}{c}{{\sc loss per query}}
        \\
        {\sc synthetic 4} & {\sc synthetic 5} & {\sc synthetic 4} & {\sc synthetic 5}
        \\
        \hline \hline
        \multicolumn{4}{c}{{\sc uniform}}
        \\
        \includegraphics[width=10em]{figures/synthetic_vs_self_4_uniform_per_iter_loss} &
        \includegraphics[width=10em]{figures/synthetic_vs_self_5_uniform_per_iter_loss} &
        \includegraphics[width=10em]{figures/synthetic_vs_self_4_uniform_per_query_loss} &
        \includegraphics[width=10em]{figures/synthetic_vs_self_5_uniform_per_query_loss}
        \\
        \hline
        \multicolumn{4}{c}{{\sc normal sparse}}
        \\
        \includegraphics[width=10em]{figures/synthetic_vs_self_4_normal_sparse_per_iter_loss} &
        \includegraphics[width=10em]{figures/synthetic_vs_self_5_normal_sparse_per_iter_loss} &
        \includegraphics[width=10em]{figures/synthetic_vs_self_4_normal_sparse_per_query_loss} &
        \includegraphics[width=10em]{figures/synthetic_vs_self_5_normal_sparse_per_query_loss}
        \\
        \hline
     \end{tabular}
    }
    \caption{\label{fig:selfcomparison} Comparison for {\sc SetMargin} with
    $k=2,3,4$.}
\end{figure}

>>>>

\paragraph{Notation.} We use boldface letters $\vx$ to indicate vectors,
uppercase letters $X$ for matrices, and calligraphic capital letters $\calX$
for sets. We frequently abbreviate the set $\{ x^i \}_{i=1}^n$ as $\{ x^i \}$
whenever the range of the index $i$ is clear from the context, and use $[n]$ as
a shorthand for the set $\{1, \ldots, n\}$. We write $\|\vx\|_1 := \sum_z |x_z|$
to indicate the $\ell_1$ vector norm, $\langle \cdot, \cdot \rangle$ to
denote the usual dot product, and $X'$ for matrix transposition.

@@

During learning, the actual weight vector $\vw$ is {\em unknown} to the
learning system, and must be estimated from pairwise choices elicited from the
user.
%\andrea{@PAOLO: qualche parola e reference per giustificare la scelta di pairwise queries?}  
Comparison queries are especially natural, asking a user if she prefers one option x to another y;
these comparisons can be extended to choice sets of more than two options \cite{viappiani2009,viappiani2010optimal}.
Comparison and choice queries are common in conjoint analysis \cite{louviere2000,toubia2004}.

@@

\section{Setwise Max-margin Learning}
\label{sec:formulation}
\paragraph{Non-linear Formulation.} We first introduce the problem
formulation as a %mixed integer 
non-linear optimization problem, and
then show how to reduce it to a mixed integer linear program.% optimization one.

The goal of our setwise max-margin approach is twofold. First, for any
given set size $k\geq 1$, we want to find a {\em set} of $k$ weight
vectors $\vw^{1}, \ldots, \vw^{k}$, chosen so that all user-provided
preferences are satisfied by the largest possible margin (i.e. all
weight vectors are consistent with respect to the user responses
$\calD$, modulo inconsistencies) and so that they are maximally
diverse.  Second, we want to construct a {\em set} of $k$
configurations $\vx^{1}, \ldots, \vx^{k}$, so that each configuration
$\vx^{i}$ is the ``best'' possible option when evaluated according to
the corresponding $\vw^{i}$ and configurations are maximally diverse
among each other. These options will be later employed to formulate
the user queries.

The first goal is achieved by translating all pairwise preferences
$\calD$ into ranking constraints: preferences of the form
$\vy^h_+ \succ \vy^h_-$ become linear inequalities of the form
$\langle \vw^i, \vy^h_+ - \vy^h_- \rangle \geq \mu$, where $\mu$ is the
{\em margin} variable (which we aim at maximizing) and $h$ ranges over
the responses.  Non-separable datasets, which occur in practice due to
occasional inconsistencies in user feedback, are handled by
introducing slack variables (whose sum we aim at minimizing). When
augmented with the slacks, the above inequalities take the form:
\begin{align*}
\langle \vw^{i}, \vy^{h}_+ - \vy^{h}_- \rangle \ge \mu - \varepsilon^{i}_h 
\end{align*}
where $\varepsilon^{i}_h$ is the penalty incurred by weight vector $\vw^{i}$
for violating the margin separation of pair $h$. Indifference preferences, i.e.
$\vy^h_1 \approx \vy^h_2$, are translated as $|\langle \vw^i, \vy^h_1 - \vy^h_2 \rangle| < \varepsilon^i_h$;
the slack increases with the difference between the estimated utility of the
two options.

@@

\begin{figure}[b]
    \centering
    {\footnotesize
    \begin{tabular}{cccc}
        \hline
        \multicolumn{2}{c}{{\sc synthetic 4}} &
        \multicolumn{2}{c}{{\sc synthetic 5}}
        \\
        {\sc loss per iter.} & {\sc loss per query} & {\sc loss per iter.} & {\sc loss per query}
        \\
        \hline \hline
        \multicolumn{4}{c}{{\sc uniform}}
        \\
        \includegraphics[width=10em]{figures/synthetic_vs_self_4_uniform_per_iter_loss} &
        \includegraphics[width=10em]{figures/synthetic_vs_self_4_uniform_per_query_loss} &
        \includegraphics[width=10em]{figures/synthetic_vs_self_5_uniform_per_iter_loss} &
        \includegraphics[width=10em]{figures/synthetic_vs_self_5_uniform_per_query_loss}
        \\
        \hline
        \multicolumn{4}{c}{{\sc normal sparse}}
        \\
        \includegraphics[width=10em]{figures/synthetic_vs_self_4_normal_sparse_per_iter_loss} &
        \includegraphics[width=10em]{figures/synthetic_vs_self_4_normal_sparse_per_query_loss} &
        \includegraphics[width=10em]{figures/synthetic_vs_self_5_normal_sparse_per_iter_loss} &
        \includegraphics[width=10em]{figures/synthetic_vs_self_5_normal_sparse_per_query_loss}
        \\
        \hline
     \end{tabular}
    }
    \caption{\label{fig:selfcomparison} Comparison for {\sc SetMargin} with
    $k=2,3,4$.}
\end{figure}
